---
title: "Tutorial_1"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Problem Sheet 1

## Problem Set A

Imported dplyr for dataframe manipulation. Used expanded grid to create the variable of all possible outcomes

```{r}
library(dplyr)

#Create vector of dice values
dice <- c(1,2,3,4,5,6)

#use expand.grid to return all possible die combinations
S <- expand.grid(dice,dice,dice)

#create new variable with sum
S <- S %>% mutate(Value=Var1+Var2+Var3)

#Calculate probability of sum of eyes=12 
prob_sum_12 <- nrow(S[S$Value == "12",])/nrow(S)

#Show value
prob_sum_12
```

Wanted to compare result with `sample` function's result. Results were similar 

```{r}

#confirm with sample
set.seed(1510)
N1 <- 1000000
diceRolls <- replicate(3, sample(dice, N1, replace = TRUE))
sum_12_2 <- sum(diceRolls[,1] + diceRolls[,2] + diceRolls[,3] == 12)
prob_sum_12_2 <- sum_12_2/N1
prob_sum_12_2

```

Next, needed to create vector of probabilities for biased die. Then renamed the columns and binded the probabilities to the variables

```{r}
#create vector of probability
biased_prob <- c(0.125,0.125,0.125,0.125,0.125,0.375)
prob_grid <- expand.grid(biased_prob,biased_prob,biased_prob)
#rename columns 
prob_grid <- prob_grid %>% rename(Prob1=Var1, Prob2=Var2, Prob3=Var3)

#join the dfs together
S <- bind_cols(S, prob_grid)
```

Ran `sample` to estimate probability of sum being 12 for biased dice

```{r}
# Calculate probability that sum is 12 
set.seed(1510)
N2 <- 1000000
BdiceRolls <- replicate(3, sample(dice, N2, replace = TRUE, prob = biased_prob))
biased_sum_12 <- sum(BdiceRolls[,1] + BdiceRolls[,2] + BdiceRolls[,3] == 12)
prob_biased_12 <- biased_sum_12/N2
prob_biased_12
```

## Problem Set B

Imported library `ggplot2` for plotting and imported data

```{r}
library(ggplot2)

#import CSVs
kenya_raw <- read.csv("P:/Adulting/Uni/Year 3/Papers/QM/Tutorial 1/kenya.csv")
sweden_raw <- read.csv("P:/Adulting/Uni/Year 3/Papers/QM/Tutorial 1/sweden.csv")
world_raw <- read.csv("P:/Adulting/Uni/Year 3/Papers/QM/Tutorial 1/world.csv")
```

Created functions to calculate ASFR, TFR, and ASDR

```{r}
#create asfr, tfr, asdr function 
asfr_tfr_asdr <- function(dataset){

  dataset <- dataset %>% mutate(asfr = births/py.women)
  dataset <- dataset %>% mutate(asdr = deaths/rowSums(dataset[,c("py.men","py.women")]))

  tfr1 <- dataset %>% slice(4:10) %>% summarise(result=sum(asfr))
  tfr2 <- dataset %>% slice(19:25) %>% summarise(result=sum(asfr))

  print(paste("The TFR for the period 1950 - 1955 is: ", (tfr1[1]*5)))
  print(paste("The TFR for the period 2005 - 2010 is: ", (tfr2[1]*5)))
  print(paste("This was a ", (format(round((((tfr2[1]-tfr1[1])/tfr1[1])*100),2))), "% change between the 2 periods"))
  return(dataset)
}
```

Ran the functions to create new dataframes to store the ASFR and ASDR. Combined the dataframes for easier plotting

```{r}
#Compute asdr, asfr, tfr for datasets
kenya_comp <- asfr_tfr_asdr(kenya_raw)
sweden_comp <- asfr_tfr_asdr(sweden_raw)
world_comp <- asfr_tfr_asdr(world_raw)
```

```{r}
#merge dfs
merge_comp <- bind_rows(kenya_comp,sweden_comp,world_comp)
merge_comp$age <- factor(merge_comp$age, levels = unique(merge_comp$age))
```

```{r}
#plot asfr
asfr_plot <- merge_comp %>%
  ggplot(aes(age, asfr, fill=period)) +
  geom_col() +
  facet_grid(~country) +
  theme(text = element_text(size=10), axis.text.x = element_text(angle=90, hjust=1),
        panel.background = element_blank()) +
  labs(title = "ASFR, segmented by Country", x = "Age Group", y = "ASFR") 
asfr_plot
```

```{r}
#plot asfr with country comparison
asfr_plot_c <- merge_comp %>%
  ggplot(aes(age, asfr, fill=country)) +
  geom_col(position="dodge") +
  facet_grid(~period) +
  theme(text = element_text(size=10), axis.text.x = element_text(angle=90, hjust=1),
        panel.background = element_blank()) +
  labs(title = "ASFR, segmented by Time Period", x = "Age Group", y = "ASFR") 
asfr_plot_c
```

```{r}
#plot asdr
asdr_plot <- merge_comp %>%
  ggplot(aes(age, asdr, fill=period)) +
  geom_col() +
  facet_grid(~country) +
  theme(text = element_text(size=10), axis.text.x = element_text(angle=90, hjust=1), 
        panel.background = element_blank()) +
  labs(title = "ASDR, segmented by Country", x = "Age Group", y = "ASDR") 
asdr_plot
```

```{r}
#plot asdr with country comparison
asdr_plot_c <- merge_comp %>%
  ggplot(aes(age, asdr, fill=country)) +
  geom_col(position="dodge") +
  facet_grid(~period) +
  theme(text = element_text(size=10), axis.text.x = element_text(angle=90, hjust=1),
        panel.background = element_blank()) +
  labs(title = "ASDR, segmented by Time Period", x = "Age Group", y = "ASDR") 
asdr_plot_c
```