---
title: "Smoking"
author: "Zachary Lim"
date: "2/27/2021"
output: html_document
---

# Introduction

Another practice document to ensure we are able to use plotly with what Elias wants us to do.

# Data Loading 

First, we load the data again, using the smoking dataset provided. 

```{r load, message=FALSE, warning=FALSE}
library(dplyr)
library(plotly)
library(tidyverse)

smokers <- read_csv("smoking.csv")
```

# Plotting 

We need to plot the association between smoking and years left to live based on age quantile. We can do this using plotly to create a more interactive plot. But first, we have to split the dataset up into different age quantiles. 

```{r break, message=FALSE, warning=FALSE}
smokers <- smokers %>% mutate(group = cut(age, breaks=c(-Inf, 50.24962, 57.61331, 63.89274, Inf),
                                           labels = c("0 - 50", "51 - 57", "58 - 63", "64 and Over")))

# Group data
df <- smokers %>%
  group_by(group, smoker) %>%
  summarise(mean = mean(years))

df$smoker <- replace(df$smoker, df$smoker==1, "Yes")
df$smoker <- replace(df$smoker, df$smoker==0, "No")

```

After we break them up, we can use plotly to plot it out using bar charts (which Elias showed was probably the clearest way to do it)

```{r, echo=FALSE, warning=FALSE, message=FALSE}
fig <- df %>% plot_ly(x=~group, y=~mean, type="bar", text=df$smoker,
                      transforms = list(list(type="groupby", 
                                             groups=df$smoker,
                                             styles=list(
                                               list(target="No", value=list(marker=list(color='blue'))),
                                               list(target="Yes", value=list(marker=list(color='darkred')))
                                             ))), hoverinfo='skip',
                      hovertemplate = paste("Smoker: %{text}",
                                            "<br>Years left:%{y: .2f}</br><extra></extra>")) %>%
  layout(barmode="group", 
         title="Comparison of Years Left to Live", 
         xaxis=list(title="Age Group Quartiles"), 
         yaxis=list(title="Years left to live"),
         showlegend=TRUE,
         legend=list(title=list(text="<b>Smoker?</b>")))

fig
```

And here's hoping it works!
